A Fredholm integral equation approach to computing Stokes eigenvalues in the plane

### 2019-03-24 doc/draft-01-stokes/01intro.tex
Updating paper

The planar incompressible Stokes equations describe
creeping flows in two dimensions.
%
Let $\Omega \subset \R^2$
be a bounded domain with $C^2$ boundary.
%
The Stokes eigenvalue problem seeks
values $k^2$ such that

@@

The eigenvalues (and eigenfunctions)
of the Stokes operator have applications in the
stability analysis of stationary solutions of the
Navier--Stokes equations \cite{osborn1976approximation}
and in the study of decaying two dimensional turbulence
\cite{schneider2008final}.
%
The eigenvalues and eigenfunctions of the Stokes
operator are the subject of intense analytical
investigation
\cite{taylor1933buckling,szego1950membranes,
  polya1951isoperimetric,bramble1963pointwise,
  ashbaugh1996fundamental,leriche2004stokes,
  kelliher2009eigenvalues,antunes2011buckling},
particularly as they relate to the eigenvalues and
eigenfunctions of the drum.

Historically, the Stokes eigenvalue problem serves as a
common model problem for numerical eigenvalue analysis
with a fourth order operator (here, the bi-Laplacian).
%
Further, numerical simulation has long played an
important role in the analyses described above --- both for
estimating the eigenvalues on domains of practical
interest and for exploring open conjectures
and generating new conjectures.

Borrowing the language of~\cite{zhao2015robust},
which concerns the Laplace or ``drum'' eigenvalue
problem,
the numerical treatment of the Stokes eigenvalue
problem can be divided into two basic approaches.
%
Type A
methods perform a direct discretization of the
Stokes operator, typically with a
finite element basis, and the eigenvalues are found
as the eigenvalues of the discrete system.
%
Type B
methods utilize a boundary integral equation
(BIE) re-formulation of the oscillatory Stokes
equation which is then discretized, so that
the eigenvalues are found by a nonlinear search
for the values of $k$ where the BIE is not invertible.
%

There is a large body of research on type A methods
for the Stokes eigenvalue problem.
%
We do not seek to review this literature here,
but point to \cite{johnson1974beam,
  rannacher1979nonconforming,
  mercier1981eigenvalue,bjorstad1999high,
  jia2009approximation,chen2006approximation,
  lovadina2009posteriori,huang2011numerical,
  carstensen2014guaranteed}
for some representative examples.

As noted in~\cite{zhao2015robust}, the type B
methods provide some advantages.
%
Because the BIE is defined on the
boundary alone, there is a reduction in the
dimension of the domain to be discretized.
%
This reduces the number of unknowns over finite
element discretizations, particularly given
that finite element methods suffer from
high-frequency ``pollution'' when $k$ is large
\cite{babuska1997pollution}.

Further, Zhao and Barnett~\cite{zhao2015robust}
show how to alleviate some of the costliness of the
nonlinear optimization introduced by a type B method.
%
The standard approach searches for ``V''-shaped minima
of the singular values of the BIE; see, for
instance, \cite{trefethen2006computed}.
%
Instead, Zhao and Barnett utilize the Fredholm
determinant (see \cref{sec:dets}) which, for certain
BIEs, is an analytic function of $k$ with roots
precisely when $k^2$ is an eigenvalue.
%
The Fredholm determinant can be estimated using
a Nystr\"{o}m discretization of the BIE
\cite{bornemann2010numerical,zhao2015robust}.
%
Then, the eigenvalues can be estimated efficiently
by using high order root finding methods applied
to the discretized determinant.

With the efficiency of the approach of
\cite{zhao2015robust} for the drum problem in mind,
we develop a type B method for the Stokes eigenvalue
problem.
%
This requires that a layer
potential representation of the solution
of \cref{eq:ostokes} be given and that the resulting BIE
is not invertible precisely when $k^2$ is an eigenvalue.
%
The first requirement is straightforward to
satisfy because the
well-known layer potential representations for the
modified Stokes equation~\cite{Pozrikidis1992,biros2002embedded,
  jiang2013second,ladyzhenskaya1969mathematical}
are directly applicable.
%
Proving the invertibility of the associated operators
away from the eigenvalues is a more involved task
and forms the bulk of the theoretical component
of this paper.

@@

While type B methods for the
related ``buckling'' eigenvalue problem
have been considered previously,
these typically relied on first-kind integral
equation formulations of the underlying PDE,
i.e. formulations in which the BIE operator is
compact \cite{kitahara2014boundary,antunes2011buckling}.
%
This is unsatisfying from a theoretical
perspective, because such operators are never invertible
(though, they are injective precisely when $k^2$ is an
eigenvalue \cite{kitahara2014boundary,antunes2011buckling})
so that the relation between the non-invertibility of the
discrete system and the eigenvalues is obscured.
%
Further,
such representations create numerical challenges because
common measures of the ``non-invertibility'' of a matrix,
like the smallest singular value or the determinant, converge
rapidly to zero for all values of $k^2$ as the boundary is
refined. Therefore, the measure of whether $k^2$ is an
approximate eigenvalue is {\em relative to the current grid}
for first kind equations.

@@

The rest of this paper proceeds as follows.
%
In \cref{sec:prelim}, we set the notation, provide some
mathematical preliminaries, and review the
properties of single and double layer potentials
for the oscillatory Stokes equations.
%
Then, in \cref{sec:analysis}, we develop the necessary
theory for proving \cref{thm:dlmain,thm:cfmain},
which show that the BIEs resulting from these
layer potential representations are not invertible
precisely when $k^2$ is an eigenvalue.
%
These theoretical developments include a detailed
discussion of the uniqueness of oscillatory Stokes
boundary value problems in exterior domains.
%
To the best of our knowledge, the invertibility
and uniqueness results are new to the literature.
%
\Cref{sec:dets} then outlines how the Fredholm determinant
can be used in the oscillatory Stokes context.
%
In \cref{sec:numerical}, we describe the numerical
methods we use to discretize the BIEs and to perform
fast determinant calculations.
%
That section also includes numerical experiments
which demonstrate some of the paper's analytical
claims as well as the effectiveness of the overall
framework.
%
Finally, we provide some concluding thoughts,
describe plans for future research,
and outline some open questions in
\cref{sec:conclusion}.

>>>>

The planar incompressible Stokes equations describe
creeping flows in two dimensions.
%
Let $\Omega \subset \R^2$
be a bounded domain with $C^2$ boundary.
%
The Stokes eigenvalue problem is to find
values $k^2$ such that

@@

The eigenvalues (and eigenfunctions)
of the Stokes operator have applications in the
stability analysis of stationary solutions of the
Navier--Stokes equations \cite{osborn1976approximation}
and in the study of decaying two dimensional turbulence
\cite{schneider2008final}.
%
The eigenvalues and eigenfunctions of the Stokes
operator are the subject of intense analytical
investigation
\cite{taylor1933buckling,szego1950membranes,
  polya1951isoperimetric,bramble1963pointwise,
  ashbaugh1996fundamental,leriche2004stokes,
  kelliher2009eigenvalues,antunes2011buckling},
particularly how they relate to the eigenvalues and
eigenfunctions of the drum.

Historically, the Stokes eigenvalue problem serves as a
common model problem for numerical eigenvalue analysis
with a fourth order operator (here, the bi-Laplacian).
%
Further, numerical simulation has long played an
important role in the analyses described above --- both for
estimating the eigenvalues on domains of practical
interest and for exploring open conjectures.

Borrowing the language of~\cite{zhao2015robust},
which concerns the Laplace or ``drum'' eigenvalue
problem,
the numerical treatment of the Stokes eigenvalue
problem can be divided into two basic approaches.
%
The first class of methods
directly discretize the
Stokes operator, typically with a
finite element basis, and the eigenvalues are found
as the eigenvalues of the discrete system.
%
The second class of methods reformulate the 
oscillatory Stokes equations as a boundary integral
equations which are discretized.
The eigenvalues are then found
by a nonlinear search for the values of $k$ where the BIE 
is not invertible.
%

There is a large body of research on the first class of methods
for the Stokes eigenvalue problem.
%
We do not seek to review this literature here,
but point to \cite{johnson1974beam,
  rannacher1979nonconforming,
  mercier1981eigenvalue,bjorstad1999high,
  jia2009approximation,chen2006approximation,
  lovadina2009posteriori,huang2011numerical,
  carstensen2014guaranteed}
for some representative examples.

As noted in~\cite{zhao2015robust}, integral equation based
methods provide several advantages.
%
Because the BIE is defined on the
boundary alone, there is a reduction in the
dimension of the domain to be discretized.
%
This reduces the number of unknowns over finite
element discretizations, particularly given
that finite element methods suffer from
high-frequency ``pollution'' when $k$ is large
\cite{babuska1997pollution}.

Further, Zhao and Barnett~\cite{zhao2015robust}
show how to alleviate some of the costliness of the
nonlinear optimization introduced by formulating the 
problem as an integral equation.
%
The standard approach searches for ``V''-shaped minima
of the singular values of the BIE; see, for
instance, \cite{trefethen2006computed}.
%
Instead, Zhao and Barnett utilize the Fredholm
determinant (see \cref{sec:dets}) which, for certain
BIEs, is an analytic function of $k$ with roots
precisely when $k^2$ is an eigenvalue.
%
The Fredholm determinant can be estimated using
a Nystr\"{o}m discretization of the BIE
\cite{bornemann2010numerical,zhao2015robust}.
%
Then, the eigenvalues can be estimated efficiently
by using high order root finding methods applied
to the discretized determinant.

With the efficiency of the approach of
\cite{zhao2015robust} for the drum problem in mind,
we develop an integral equation based method for the Stokes eigenvalue
problem.
%
This requires that a layer
potential representation of the solution
of \cref{eq:ostokes} be given and that the resulting BIE
is not invertible precisely when $k^2$ is an eigenvalue.
%
The first requirement is straightforward to
satisfy because the
well-known layer potential representations for the
modified Stokes equation~\cite{Pozrikidis1992,biros2002embedded,
  jiang2013second,ladyzhenskaya1969mathematical}
are directly applicable.
%
Proving the invertibility of the associated operators
away from the eigenvalues is a more involved task
and forms the bulk of the theoretical component
of this paper.

@@

While integral equation based methods for the
related ``buckling'' eigenvalue problem
have been considered previously,
these typically relied on first-kind integral
equation formulations of the underlying PDE,
i.e. formulations in which the BIE operator is
compact \cite{kitahara2014boundary,antunes2011buckling}.
%
This is unsatisfying from a theoretical
perspective, because such operators are never \note{Travis: why is this the case?} 
invertible
(though, they are not injective precisely when $k^2$ is an
eigenvalue \cite{kitahara2014boundary,antunes2011buckling})
so that the relation between the non-invertibility of the
discrete system and the eigenvalues is obscured.
%
Further,
such representations create numerical challenges because
common measures of the ``non-invertibility'' of a matrix,
like the smallest singular value or the determinant, converge
rapidly to zero for all values of $k^2$ as the boundary is
refined. Therefore, the measure of whether $k^2$ is an
approximate eigenvalue is {\em relative to the current grid}
for first kind equations.

@@

The rest of this paper proceeds as follows.
%
In \cref{sec:prelim}, we set the notation, provide some
mathematical preliminaries, and review the
properties of single and double layer potentials
for the oscillatory Stokes equations.
%
Then, in \cref{sec:analysis}, we develop the necessary
theory for proving the main results of this work 
(\cref{thm:dlmain,thm:cfmain}),
which show that the BIEs resulting from these
layer potential representations are not invertible
precisely when $k^2$ is an eigenvalue.
%
These theoretical developments include a detailed
discussion of the uniqueness of oscillatory Stokes
boundary value problems in exterior domains.
%
To the best of our knowledge, the invertibility
and uniqueness results are new to the literature.
%
\Cref{sec:dets} then outlines how the Fredholm determinant
can be used in the oscillatory Stokes context.
%
In \cref{sec:numerical}, we describe the numerical
methods we use to discretize the BIEs and to perform
fast determinant calculations.
%
That section also includes numerical experiments
which demonstrate some of the paper's analytical
claims as well as the effectiveness of the overall
framework.
%
Finally, we provide some concluding thoughts,
describe plans for future research,
and outline some open questions in
\cref{sec:conclusion}.

### 2019-03-24 doc/draft-01-stokes/02prelim.tex
Updating paper

\section{Mathematical Preliminaries}
\label{sec:prelim}
In this paper, vector-valued
quantities are denoted by bold, lower-case letters
(e.g. $\bh$), while tensor-valued quantities are bold
and upper-case (e.g. $\mathbf{T}$). 
Subscript indices of non-bold characters (e.g. $h_j$ or $T_{jkl}$)
are used to denote the entries within a vector ($\bh$) or tensor ($\bT$).
We use the standard Einstein summation convention; in other words, 
there is an implied sum taken over the repeated indices of 
any term (e.g. the symbol $a_{j} b_{j}$ is used to represent the sum
$\sum_{j} a_{j} b_{j}$).
If $\xx = (x_1,x_2)^\intercal$, then $\xx^\bot = (-x_2,x_1)^\intercal$.
Similarly, $\nabla^\bot = (-\partial_{x_2},\partial_{x_1})^\intercal$.
Upper-case script characters (e.g. $\mathcal{K}$) are reserved for
operators on Banach spaces, with $\mathcal{I}$ denoting the
identity.

@@

We seek solutions of \cref{eq:ostokes} in the space
\note{make sure this is the correct space}
\begin{equation}
  A(\Omega) = \{ (\bu,p) \textrm{ s.t. } \bu \in
  \left ( C^2(\Omega)\times C^2(\Omega) \right ) \cap
  \left ( C(\bar{\Omega}) \times C(\bar{\Omega}) \right ) \, , \,
  p \in C^1(\Omega) \times C(\bar{\Omega})\} \; ,
\end{equation}
where $\Omega$ is an open domain and $\bar{\Omega}$ denotes
its closure.

@@

Let $\mathcal{L}_x$ denote a linear differential operator. A fundamental
solution $G(\xx,\yy)$ of $\mathcal{L}_x$ satisfies the equation
$\mathcal{L}_x G(\xx,\yy) = \delta_y(\xx)$ in the distributional sense, i.e.
for sufficiently smooth $f$
\begin{equation}
  \mathcal{L}_x \int_{\R^2} G(\xx,\yy) f(\yy) \, dy = f(\xx) \; .
  \nonumber
\end{equation}
We consider here
``free-space'' Green's functions, i.e. fundamental solutions which satisfy
radiation conditions as $|\xx-\yy| \to \infty$.
The Green's function of the oscillatory biharmonic equation,
\begin{equation}
  \Delta ( \Delta + k^2 ) u = 0 \; , \label{eq:obiharm} \nonumber
\end{equation}
is given by 
\begin{equation}
  \Gbh(\xx,\yy) = \frac{1}{k^2}
  \left (\frac{1}{2\pi} \log |\xx-\yy| +
  \frac{i}{4} H_0^{(1)}(k|\xx-\yy|) \right ) \, ,
  \label{eq:Gbh}
\end{equation}
where $k$ is the Helmholtz paramater in the oscillatory biharmonic equation,
and $H_{0}^{1}(r)$ is the Hankel function of the first kind of order zero.
Note that this is a scaled difference of the Green's function for
Laplace, i.e.

@@

We first recall some definitions.
Let $X$ and $Y$ be Banach spaces with a non-degenerate
bilinear form $\langle \cdot ,\cdot \rangle: X\times Y \to \C$.
\begin{itemize}
\item Two operators $\cA:X\to X$ and $\cB:Y\to Y$ are
adjoint operators if
$\langle A \phi,\psi \rangle = \langle \phi, B\psi \rangle$
for every $\phi \in X$ and $\psi \in Y$.
\item For an operator $\cM:X\to X$, we can define the range
  $R(\cM)$ as the set $\{\phi \in X: \exists \phi_0 \textrm{ with }
  \cM\phi_0 = \phi\}$ and the null space $N(\cM)$ as the
  set $\{\phi \in X: \cM \phi = 0 \}$.
\item An operator $\cA$ is said to be compact if
  $\overbar{\cA V}$ is a compact set for any

\item Given a subspace $V\subset X$,
  we can define the subspace $V^\perp\subset Y$ as the
  set $V^\perp = \{ \psi \in Y: \langle \phi,\psi \rangle = 0
  \textrm{ for each } \phi \in V \}$, with the analogous
  definition for subspaces of $Y$.
\end{itemize}

@@

\subsubsection{Oscillatory Stokeslets and stresslets}

In this section, we review the derivation of the
oscillatory Stokeslet, which is the appropriate setting for the Stokes
eigenvalue problem.
 Consider the solution of
\cref{eq:ostokes} where a $\delta$-mass
centered at $\yy$ with strength $\ff$
has been added to the right-hand side of \cref{eq:ostokes}, i.e.

@@

\begin{align}
  T_{ij\ell} &= - \partial_{x_j} \Glap \delta_{i\ell}
  + \partial_{x_\ell} \left ( -\Delta \Gbh \delta_{ij} +
  \partial_{x_i} \left(\partial_{x_j} \Gbh \right) \right)
  \nonumber \\
  & \qquad+ \partial_{x_i} \left ( -\Delta \Gbh \delta_{\ell j} +
  \partial_{x_\ell} \left(\partial_{x_j} \Gbh \right) \right)
  \; . \label{eq:ostress} \nonumber
\end{align}
Let $u_i = G_{ij} f_j$ and $p = \partial_{x_i} \Glap f_i$ be a
solution of the Stokes equations induced by a Stokeslet.
Then the corresponding stress tensor is given by
$\sigma_{i\ell} = T_{ij\ell} f_j$.

@@

  \begin{align}
    \lim_{h \downarrow 0} \bsigma_\bS[\bmu](\xx_0 \pm h\bnu(\xx_0)) \cdot \bnu(\xx_0)
    &= \mp \frac{1}{2} \bmu(\xx_0) + \cDt[\bmu](\xx_0) \\
    \lim_{h \downarrow 0} \bD[\bmu](\xx_0 \pm h\bnu(\xx_0)) 
    &= \pm \frac{1}{2} \bmu(\xx_0) + \cD[\bmu](\xx_0)    \; .
  \end{align}
\end{lem}

@@

In the following theorem, we sketch the proof of the equivalent of the
Green's identity for oscillatory Stokes, which is well-known
in the Stokes and modified Stokes settings
\cite{Pozrikidis1992,biros2002embedded,ladyzhenskaya1969mathematical}.

>>>>

\section{Mathematical Preliminaries}
\label{sec:prelim}
In this paper, vector-valued
quantities are denoted by bold, lower-case letters
(e.g. $\bh$), while tensor-valued quantities are bold
and upper-case (e.g. $\mathbf{T}$). 
Subscript indices of non-bold characters (e.g. $h_j$ or $T_{jkl}$)
are used to denote the entries within a vector ($\bh$) or tensor ($\bT$).
We use the standard Einstein summation convention; in other words, 
there is an implied sum taken over the repeated indices of 
any term (e.g. the symbol $a_{j} b_{j}$ is used to represent the sum
$\sum_{j} a_{j} b_{j}$).
If $\xx = (x_1,x_2)^\intercal$, then $\xx^\bot = (-x_2,x_1)^\intercal$.
Similarly, $\nabla^\bot = (-\partial_{x_2},\partial_{x_1})^\intercal$.
Upper-case script characters (e.g. $\mathcal{K}$) are reserved for
operators on Banach spaces, with $\mathcal{I}$ denoting the
identity. Given a set $X$, we denote the closure of $X$
by $\overline{X}$.

@@

We seek solutions of \cref{eq:ostokes} in the space
\note{make sure this is the correct space: (MR - looks good to me)}
\begin{equation}
  A(\Omega) = \{ (\bu,p) \textrm{ s.t. } \bu \in
  \left ( C^2(\Omega)\times C^2(\Omega) \right ) \cap
  \left ( C(\bar{\Omega}) \times C(\bar{\Omega}) \right ) \, , \,
  p \in C^1(\Omega) \times C(\bar{\Omega})\} \; ,
\end{equation}
where $\Omega$ is an open domain.

@@

Let $\mathcal{L}_x$ denote a linear differential operator. A fundamental
solution $G(\xx,\yy)$ of $\mathcal{L}_x$ satisfies the equation
$\mathcal{L}_x G(\xx,\yy) = \delta_y(\xx)$ in the distributional sense, i.e.
for sufficiently smooth $f$
\begin{equation}
  \mathcal{L}_x \int_{\R^2} G(\xx,\yy) f(\yy) \, d \by = f(\xx) \; .
  \nonumber
\end{equation}
We consider here
free-space Green's functions, i.e. fundamental solutions which satisfy
radiation conditions as $|\xx-\yy| \to \infty$.
The Green's function of the oscillatory biharmonic equation,
\begin{equation}
  \Delta ( \Delta + k^2 ) u = 0 \; , \label{eq:obiharm} \nonumber
\end{equation}
is given by 
\begin{equation}
  \Gbh(\xx,\yy) = \frac{1}{k^2}
  \left (\frac{1}{2\pi} \log |\xx-\yy| +
  \frac{i}{4} H_0^{(1)}(k|\xx-\yy|) \right ) \, ,
  \label{eq:Gbh}
\end{equation}
where $k$ is the Helmholtz paramater in the oscillatory biharmonic equation,
and $H_{0}^{1}(r)$ is the Hankel function of the first kind of order zero.
Note that this is a scaled difference of the Green's function for
Laplace, i.e.

@@

We first recall some definitions.
Let $X$ and $Y$ be Banach spaces with a non-degenerate
bilinear form $\langle \cdot ,\cdot \rangle: X\times Y \to \C$.
\begin{itemize}
\item Two operators $\cA:X\to X$ and $\cB:Y\to Y$ are
adjoint operators if
$\langle A \phi,\psi \rangle = \langle \phi, B\psi \rangle$
for every $\phi \in X$ and $\psi \in Y$.
\item For an operator $\cM:X\to X$, we can define the range
  $R(\cM)$ as the set $\{\phi \in X: \exists \phi_0 \textrm{ with }
  \cM\phi_0 = \phi\}$ and the null space $N(\cM)$ as the
  set $\{\phi \in X: \cM \phi = 0 \}$.
\item An operator $\cA$ is said to be compact if
  $\overline{\cA V}$ is a compact set for any
  bounded subset $V\subset X$.
\item Given a subspace $V\subset X$,
  we can define the subspace $V^\perp\subset Y$ as the
  set $V^\perp = \{ \psi \in Y: \langle \phi,\psi \rangle = 0
  \textrm{ for each } \phi \in V \}$, with the analogous
  definition for subspaces of $Y$.
\end{itemize}

@@

\subsubsection{Oscillatory Stokeslets and stresslets}
Consider the solution of
\cref{eq:ostokes} where a $\delta$-mass
centered at $\yy$ with strength $\ff$
has been added to the right-hand side of \cref{eq:ostokes}, i.e.

@@

\begin{align}
  T_{ij\ell} &= - \partial_{x_j} \Glap \delta_{i\ell}
  + \partial_{x_\ell} \left ( -\Delta \Gbh \delta_{ij} +
  \partial_{x_i} \left(\partial_{x_j} \Gbh \right) \right)
  \nonumber \\
  & \qquad+ \partial_{x_i} \left ( -\Delta \Gbh \delta_{\ell j} +
  \partial_{x_\ell} \left(\partial_{x_j} \Gbh \right) \right)
  \; . \label{eq:ostress} 
\end{align}
Let $u_i = G_{ij} f_j$ and $p = \partial_{x_i} \Glap f_i$ be a
solution of the Stokes equations induced by a Stokeslet.
Then the corresponding stress tensor is given by
$\sigma_{i\ell} = T_{ij\ell} f_j$.

@@

  \begin{align}
    \lim_{h \downarrow 0^{+}} \bsigma_\bS[\bmu](\xx_0 \pm h\bnu(\xx_0)) \cdot \bnu(\xx_0)
    &= \mp \frac{1}{2} \bmu(\xx_0) + \cDt[\bmu](\xx_0) \\
    \lim_{h \downarrow 0^{+}} \bD[\bmu](\xx_0 \pm h\bnu(\xx_0)) 
    &= \pm \frac{1}{2} \bmu(\xx_0) + \cD[\bmu](\xx_0)    \; .
  \end{align}
\end{lem}

@@

In the following theorem, we sketch the proof of the equivalent of the
Green's identity for oscillatory Stokes setting, which is well-known
in the Stokes and modified Stokes settings
\cite{Pozrikidis1992,biros2002embedded,ladyzhenskaya1969mathematical}.
